---
title: "Final cross-dataset analysis: Kather-5K and CRC-VAL-HE-7K"
author: "Haochen Tan"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
params:
  bootstrap_iterations: 5000
  bootstrap_seed: 2027
  denominator_threshold: 0.25
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(scales)

if (!exists("params")) {
  params <- list(
    bootstrap_iterations = 5000,
    bootstrap_seed = 2027,
    denominator_threshold = 0.25
  )
}

project_root <- normalizePath(file.path(getwd(), ".."), mustWork = TRUE)
external_root <- file.path(
  project_root, "artifacts", "crc_val_external", "common_seven"
)
kather_root <- file.path(project_root, "artifacts", "kather5k_jpi_revision")
output_dir <- file.path(external_root, "final_reporting")
table_dir <- file.path(output_dir, "tables")
figure_dir <- file.path(project_root, "analysis", "manuscript_figures")
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

model_order <- c("ResNet18", "DINOv2", "UNI")
model_colors <- c(
  ResNet18 = "#3266A8",
  DINOv2 = "#D17A22",
  UNI = "#23866B"
)
primary_methods <- c(
  ResNet18 = "gradcam",
  DINOv2 = "gradient_attention_rollout",
  UNI = "gradient_attention_rollout"
)

theme_manuscript <- function() {
  theme_minimal(base_size = 10.5) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title = element_text(color = "#202020"),
      axis.text = element_text(color = "#303030"),
      plot.title = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(color = "#505050", size = 9),
      legend.position = "bottom",
      legend.title = element_blank(),
      strip.text = element_text(face = "bold"),
      plot.margin = margin(6, 8, 6, 6)
    )
}

save_figure <- function(plot, stem, width, height) {
  ggsave(
    file.path(figure_dir, paste0(stem, ".pdf")), plot,
    width = width, height = height, units = "in", device = grDevices::pdf
  )
  ggsave(
    file.path(figure_dir, paste0(stem, ".png")), plot,
    width = width, height = height, units = "in", dpi = 320,
    bg = "white"
  )
}
```

# Locked analysis

This report is a CPU-only post-processing analysis. It does not retrain a model,
change the frozen 98-image external attribution cohort, or recalculate attribution
maps. The only amendment is the prespecified denominator sensitivity rule:

\[
  |z_y(x)| < 0.25.
\]

For a paired cross-model deletion comparison, an image-seed set is retained only
when all three models satisfy the denominator rule. This preserves pairing while
avoiding model-specific removal of observations after inspecting ratios.

```{r load-data}
metrics <- read.csv(
  file.path(external_root, "faithfulness", "external_faithfulness_metrics.csv"),
  stringsAsFactors = FALSE
)
classification <- read.csv(
  file.path(external_root, "classification", "classification_per_seed.csv"),
  stringsAsFactors = FALSE
)
classification_class <- read.csv(
  file.path(external_root, "classification", "classification_per_class.csv"),
  stringsAsFactors = FALSE
)
kather_classification <- read.csv(
  file.path(kather_root, "classification", "classification_per_seed.csv"),
  stringsAsFactors = FALSE
)
kather_faithfulness <- read.csv(
  file.path(kather_root, "statistics", "architecture_explainer_summary.csv"),
  stringsAsFactors = FALSE
)

metrics <- metrics %>%
  filter(
    perturbation == "normalized_zero",
    grid_label == "common_14",
    method == unname(primary_methods[model])
  ) %>%
  mutate(
    model = factor(model, levels = model_order),
    top_beats_random_target_logit =
      as.numeric(
        tolower(as.character(top_beats_random_target_logit)) %in%
          c("true", "1")
      )
  )

external_primary <- metrics %>%
  filter(analysis_family == "primary_jointly_correct_true_class")
external_all <- metrics %>%
  filter(analysis_family == "all_image_true_class_sensitivity")
```

# Denominator-filtered class-complete analysis

```{r matched-filter}
flag_audit <- external_all %>%
  mutate(
    denominator_flag = abs(unperturbed_target_logit) <
      params$denominator_threshold
  ) %>%
  select(
    cohort_id, seed, model, class_name, unperturbed_target_logit,
    denominator_flag
  )

matched_sets <- flag_audit %>%
  group_by(cohort_id, seed) %>%
  summarise(
    any_denominator_flag = any(denominator_flag),
    flagged_models = paste(as.character(model[denominator_flag]), collapse = ";"),
    .groups = "drop"
  )

exclusion_audit <- flag_audit %>%
  filter(denominator_flag) %>%
  arrange(model, cohort_id, seed)

external_all_filtered <- external_all %>%
  inner_join(
    matched_sets %>% filter(!any_denominator_flag) %>%
      select(cohort_id, seed),
    by = c("cohort_id", "seed")
  )

stopifnot(n_distinct(external_all_filtered$model) == 3)
stopifnot(
  all(
    external_all_filtered %>%
      count(cohort_id, seed) %>%
      pull(n) == 3
  )
)

write.csv(
  exclusion_audit,
  file.path(table_dir, "external_near_zero_logit_exclusion_audit.csv"),
  row.names = FALSE
)
write.csv(
  matched_sets,
  file.path(table_dir, "external_matched_image_seed_retention.csv"),
  row.names = FALSE
)

data.frame(
  denominator_threshold = params$denominator_threshold,
  external_images = n_distinct(external_all$cohort_id),
  original_image_seed_sets = n_distinct(
    interaction(external_all$cohort_id, external_all$seed)
  ),
  flagged_model_rows = nrow(exclusion_audit),
  flagged_resnet18_rows = sum(exclusion_audit$model == "ResNet18"),
  flagged_dinov2_rows = sum(exclusion_audit$model == "DINOv2"),
  flagged_uni_rows = sum(exclusion_audit$model == "UNI"),
  excluded_paired_image_seed_sets = sum(matched_sets$any_denominator_flag),
  retained_paired_image_seed_sets = sum(!matched_sets$any_denominator_flag)
)
```

The rule identifies eight ResNet18 rows and two UNI rows. No DINOv2 row is
flagged. Removing the corresponding image-seed sets from all three models retains
284 of 294 paired image-seed sets and all 98 unique images.

## Paired inference

```{r inference-functions}
paired_bootstrap_ci <- function(difference, iterations, seed) {
  set.seed(seed)
  boot <- replicate(
    iterations,
    mean(sample(difference, length(difference), replace = TRUE))
  )
  unname(quantile(boot, c(0.025, 0.975), na.rm = TRUE))
}

paired_tests <- function(data, metric_names, family_name) {
  image_values <- data %>%
    group_by(cohort_id, class_name, model) %>%
    summarise(across(all_of(metric_names), ~mean(.x, na.rm = TRUE)),
              .groups = "drop")

  pairs <- list(
    c("UNI", "ResNet18"),
    c("UNI", "DINOv2"),
    c("DINOv2", "ResNet18")
  )
  rows <- list()
  index <- 1L
  for (metric_name in metric_names) {
    wide <- image_values %>%
      select(cohort_id, model, all_of(metric_name)) %>%
      pivot_wider(names_from = model, values_from = all_of(metric_name))
    for (pair in pairs) {
      difference <- wide[[pair[[1]]]] - wide[[pair[[2]]]]
      difference <- difference[is.finite(difference)]
      ci <- paired_bootstrap_ci(
        difference,
        params$bootstrap_iterations,
        params$bootstrap_seed + index
      )
      if (length(difference) < 2L || all(abs(difference) < 1e-12)) {
        test_statistic <- 0
        test_p <- 1
      } else {
        test <- suppressWarnings(
          wilcox.test(difference, mu = 0, paired = FALSE,
                      exact = FALSE, correct = FALSE)
        )
        test_statistic <- unname(test$statistic)
        test_p <- test$p.value
      }
      rows[[index]] <- data.frame(
        analysis_family = family_name,
        metric = metric_name,
        model_a = pair[[1]],
        model_b = pair[[2]],
        difference_definition = "model_a_minus_model_b",
        paired_images = length(difference),
        mean_difference = mean(difference),
        median_difference = median(difference),
        bootstrap_ci_low = ci[[1]],
        bootstrap_ci_high = ci[[2]],
        bootstrap_inference_level = "paired_tile_bootstrap_exploratory",
        wilcoxon_statistic = test_statistic,
        wilcoxon_p = test_p,
        stringsAsFactors = FALSE
      )
      index <- index + 1L
    }
  }
  bind_rows(rows) %>%
    group_by(metric) %>%
    mutate(wilcoxon_p_holm = p.adjust(wilcoxon_p, method = "holm")) %>%
    ungroup()
}
```

```{r filtered-tests}
deletion_metrics <- c(
  "top_minus_random_relative_target_logit_reduction_auc",
  "top_beats_random_target_logit",
  "top_minus_random_5_relative_target_logit_reduction",
  "top_minus_random_10_relative_target_logit_reduction",
  "top_minus_random_20_relative_target_logit_reduction",
  "top_minus_random_30_relative_target_logit_reduction"
)

filtered_tests <- paired_tests(
  external_all_filtered,
  deletion_metrics,
  "all_image_true_class_denominator_filtered"
)
write.csv(
  filtered_tests,
  file.path(table_dir, "external_filtered_paired_model_tests.csv"),
  row.names = FALSE
)

filtered_tests %>%
  filter(metric %in% c(
    "top_minus_random_relative_target_logit_reduction_auc",
    "top_beats_random_target_logit"
  )) %>%
  select(
    metric, model_a, model_b, paired_images, mean_difference,
    bootstrap_ci_low, bootstrap_ci_high, wilcoxon_p_holm
  )
```

## Filtered descriptive summaries

```{r filtered-summaries}
summary_metrics <- c(
  "attribution_occlusion_spearman",
  "top_minus_random_relative_target_logit_reduction_auc",
  "top_beats_random_target_logit"
)

filtered_overall <- external_all_filtered %>%
  group_by(model) %>%
  summarise(
    image_seed_rows = n(),
    images = n_distinct(cohort_id),
    across(
      all_of(summary_metrics),
      list(mean = ~mean(.x, na.rm = TRUE),
           median = ~median(.x, na.rm = TRUE),
           sd = ~sd(.x, na.rm = TRUE))
    ),
    .groups = "drop"
  )

filtered_by_class <- external_all_filtered %>%
  group_by(model, class_name) %>%
  summarise(
    image_seed_rows = n(),
    images = n_distinct(cohort_id),
    across(
      all_of(summary_metrics),
      list(mean = ~mean(.x, na.rm = TRUE),
           median = ~median(.x, na.rm = TRUE),
           sd = ~sd(.x, na.rm = TRUE))
    ),
    .groups = "drop"
  )

write.csv(
  filtered_overall,
  file.path(table_dir, "external_filtered_class_complete_overall.csv"),
  row.names = FALSE
)
write.csv(
  filtered_by_class,
  file.path(table_dir, "external_filtered_class_complete_by_class.csv"),
  row.names = FALSE
)

filtered_overall
```

# Publication figures

## External classification

```{r external-classification-figure, fig.width=10.5, fig.height=4.2}
classification$model <- factor(classification$model, levels = model_order)
classification_class$model <- factor(
  classification_class$model, levels = model_order
)

class_labels <- c(
  adipose = "Adipose", background = "Background",
  debris_mucus = "Debris/mucus", lymphocytes = "Lymphocytes",
  normal_mucosa = "Normal mucosa", stroma = "Stroma", tumor = "Tumor"
)

classification_long <- classification %>%
  pivot_longer(
    c(accuracy, balanced_accuracy, macro_f1),
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      accuracy = "Accuracy",
      balanced_accuracy = "Balanced accuracy",
      macro_f1 = "Macro-F1"
    )
  )

g_classification <- ggplot(
  classification_long,
  aes(model, value, color = model, group = seed)
) +
  geom_line(color = "#B8B8B8", linewidth = 0.35) +
  geom_point(size = 2.1, alpha = 0.9) +
  stat_summary(
    aes(group = model), fun = mean, geom = "point",
    shape = 23, size = 3.2, fill = "white", color = "black"
  ) +
  facet_wrap(~metric, nrow = 1) +
  scale_color_manual(values = model_colors) +
  scale_y_continuous(limits = c(0.58, 0.86), breaks = seq(0.6, 0.85, 0.05)) +
  labs(
    title = "A. Coarse-label external classification",
    subtitle = "Kather-trained models evaluated on 6,588 CRC-VAL tiles; three training seeds",
    x = NULL, y = NULL
  ) +
  theme_manuscript() +
  theme(legend.position = "none")

class_heatmap <- classification_class %>%
  group_by(model, class_name) %>%
  summarise(recall = mean(recall), .groups = "drop") %>%
  mutate(
    class_name = factor(
      class_name,
      levels = rev(names(class_labels)),
      labels = rev(unname(class_labels))
    )
  )

g_class_heatmap <- ggplot(class_heatmap, aes(model, class_name, fill = recall)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", recall)), size = 3.0) +
  scale_fill_gradientn(
    colors = c("#F7F7F7", "#F3C567", "#23866B"),
    limits = c(0, 1), labels = percent_format(accuracy = 1)
  ) +
  labs(
    title = "B. Tissue-specific recall",
    subtitle = "Mean across three seeds",
    x = NULL, y = NULL, fill = "Recall"
  ) +
  theme_manuscript() +
  theme(panel.grid = element_blank(), legend.position = "right")

figure_external_classification <-
  g_classification + g_class_heatmap + plot_layout(widths = c(1.25, 0.9))
save_figure(
  figure_external_classification,
  "figure_13_external_classification",
  width = 11.2, height = 4.25
)
figure_external_classification
```

## Cross-dataset faithfulness

```{r cross-dataset-faithfulness-figure, fig.width=10.5, fig.height=4.4}
kather_primary <- kather_faithfulness %>%
  filter(
    cohort_name == "selected_faithfulness",
    analysis_family == "primary_jointly_correct_true_class",
    perturbation == "normalized_zero",
    grid_label == "common_14",
    model %in% model_order
  ) %>%
  transmute(
    scope = "Kather conditional\n187 images, 3 seeds",
    model,
    attribution_occlusion_spearman = mean_spearman,
    top_minus_random_relative_target_logit_reduction_auc =
      mean_scale_free_logit_auc,
    top_beats_random_target_logit = top_beats_random_fraction
  )

external_primary_summary <- external_primary %>%
  group_by(model) %>%
  summarise(across(all_of(summary_metrics), ~mean(.x, na.rm = TRUE)),
            .groups = "drop") %>%
  mutate(scope = "CRC conditional\n48 images, 3 seeds")

external_filtered_summary <- external_all_filtered %>%
  group_by(model) %>%
  summarise(across(all_of(summary_metrics), ~mean(.x, na.rm = TRUE)),
            .groups = "drop") %>%
  mutate(scope = "CRC class-complete\n98 images, filtered")

cross_dataset_summary <- bind_rows(
  kather_primary,
  external_primary_summary,
  external_filtered_summary
) %>%
  mutate(
    model = factor(model, levels = model_order),
    scope = factor(
      scope,
      levels = c(
        "Kather conditional\n187 images, 3 seeds",
        "CRC conditional\n48 images, 3 seeds",
        "CRC class-complete\n98 images, filtered"
      )
    )
  )

write.csv(
  cross_dataset_summary,
  file.path(table_dir, "cross_dataset_faithfulness_summary.csv"),
  row.names = FALSE
)

cross_dataset_long <- cross_dataset_summary %>%
  pivot_longer(
    all_of(summary_metrics), names_to = "metric", values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      attribution_occlusion_spearman = "Attribution-occlusion Spearman",
      top_minus_random_relative_target_logit_reduction_auc =
        "Top - random relative logit AUC",
      top_beats_random_target_logit = "Top deletion beats random"
    )
  )

figure_cross_dataset <- ggplot(
  cross_dataset_long,
  aes(scope, value, color = model, group = model)
) +
  geom_hline(yintercept = 0, color = "#8A8A8A", linewidth = 0.35) +
  geom_line(linewidth = 0.7, alpha = 0.75) +
  geom_point(size = 2.7) +
  facet_wrap(~metric, scales = "free_y", nrow = 1) +
  scale_color_manual(values = model_colors) +
  labs(
    title = "Faithfulness rankings change across dataset and estimand",
    subtitle = "Positive relative deletion AUC means top-ranked deletion is more damaging than random",
    x = NULL, y = NULL
  ) +
  theme_manuscript() +
  theme(axis.text.x = element_text(angle = 18, hjust = 1))

save_figure(
  figure_cross_dataset,
  "figure_14_cross_dataset_faithfulness",
  width = 11.0, height = 4.35
)
figure_cross_dataset
```

## External tissue heterogeneity

```{r external-heterogeneity-figure, fig.width=9.2, fig.height=5.2}
heterogeneity <- external_all_filtered %>%
  group_by(model, class_name) %>%
  summarise(
    `Attribution-occlusion Spearman` =
      mean(attribution_occlusion_spearman, na.rm = TRUE),
    `Top - random relative logit AUC` =
      mean(top_minus_random_relative_target_logit_reduction_auc),
    .groups = "drop"
  ) %>%
  pivot_longer(
    c(`Attribution-occlusion Spearman`, `Top - random relative logit AUC`),
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(
    model = factor(model, levels = model_order),
    class_name = factor(
      class_name,
      levels = names(class_labels),
      labels = unname(class_labels)
    )
  )

figure_heterogeneity <- ggplot(
  heterogeneity,
  aes(class_name, model, fill = value)
) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", value)), size = 3.0) +
  facet_wrap(~metric, ncol = 1, scales = "free") +
  scale_fill_gradient2(
    low = "#B55245", mid = "#F7F7F7", high = "#23866B",
    midpoint = 0
  ) +
  labs(
    title = "External attribution faithfulness is tissue dependent",
    subtitle = "Class-complete true-class sensitivity; denominator-filtered matched image-seed sets",
    x = NULL, y = NULL, fill = "Mean"
  ) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = 28, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "right"
  )

save_figure(
  figure_heterogeneity,
  "figure_15_external_tissue_heterogeneity",
  width = 9.4, height = 5.5
)
figure_heterogeneity
```

# Final reporting table

```{r final-table}
classification_summary <- classification %>%
  group_by(model) %>%
  summarise(
    accuracy_mean = mean(accuracy),
    accuracy_sd = sd(accuracy),
    balanced_accuracy_mean = mean(balanced_accuracy),
    macro_f1_mean = mean(macro_f1),
    .groups = "drop"
  )

external_reporting_table <- classification_summary %>%
  left_join(
    filtered_overall %>%
      select(
        model,
        spearman_mean = attribution_occlusion_spearman_mean,
        relative_logit_auc_mean =
          top_minus_random_relative_target_logit_reduction_auc_mean,
        top_beats_random_fraction = top_beats_random_target_logit_mean
      ),
    by = "model"
  ) %>%
  arrange(factor(model, levels = model_order))

write.csv(
  external_reporting_table,
  file.path(table_dir, "external_final_reporting_table.csv"),
  row.names = FALSE
)
external_reporting_table
```

# Reproducibility manifest

```{r manifest}
manifest <- data.frame(
  item = c(
    "source_metrics",
    "source_classification",
    "cohort_manifest",
    "denominator_threshold",
    "bootstrap_iterations",
    "bootstrap_seed",
    "inference_level",
    "patient_metadata_available",
    "models_retrained"
  ),
  value = c(
    "faithfulness/external_faithfulness_metrics.csv",
    "classification/classification_per_seed.csv",
    "preflight/common_seven_attribution_cohort_tile_exploratory.csv",
    as.character(params$denominator_threshold),
    as.character(params$bootstrap_iterations),
    as.character(params$bootstrap_seed),
    "paired tile bootstrap exploratory",
    "false",
    "false"
  )
)
write.csv(
  manifest,
  file.path(output_dir, "final_reporting_manifest.csv"),
  row.names = FALSE
)
manifest
```

# Interpretation lock

The filtered analysis supports a deliberately non-universal conclusion. DINOv2 has
the strongest CRC-VAL classification and patchwise attribution-occlusion agreement.
UNI has the most favorable numerical top-versus-random deletion behavior. The two
Transformer pipelines are complementary rather than reducible to a single external
winner. Highlighted patches are regions contributing to a model score under the
defined perturbation; they are not biological causes.
